---
title: "p8130_hw5_xx2485"
author: "Xiaoni Xu"
date: "2024-12-15"
output: html_document
---

```{r}
library(faraway)
```

## a)

Provide descriptive statistics for all variables of interest (continuous and categorical) –
no test required.

```{r}
data(state)

state_data <- as.data.frame(state.x77)

str(state_data)

# Generate summary statistics for all variables
summary(state_data)
```

## b)

Examine exploratory plots, e.g., scatter plots, histograms, box-plots to get a sense of
the data and possible variable transformations. (Be selective! Even if you create 20
plots, you don’t want to show them all). If you find a transformation to be necessary or
recommended, perform the transformation and use it through the rest of the problem.

```{r}

```

## c)
Use automatic procedures to find a ‘best subset’ of the full model. Present the results
and comment on the following:
• Do the procedures generate the same model?
• Are any variables a close call? What was your decision: keep or discard? Provide
arguments for your choice. (Note: this question might have more or less relevance
depending on the ‘subset’ you choose).
• Is there any association between ‘Illiteracy’ and ‘HS graduation rate’? Does your ‘subset’
contain both?
```{r}

```

## d)
Use criterion-based procedures to guide your selection of the ‘best subset’. Summarize
your results (tabular or graphical).
```{r}

```


## e)
Use the LASSO method to perform variable selection. Make sure you choose the “best
lambda” to use and show how you determined this.

```{r}

```

## f)

Compare the ‘subsets’ from parts c, d, and e and recommend a ‘final’ model. Using this
‘final’ model do the following:
• Check the model assumptions.
• Test the model predictive ability using a 10-fold cross-validation.

```{r}

```

## g)
In a paragraph, summarize your findings to address the primary question posed by the
investigator (that has limited statistical knowledge).

